---
title: "Homework 7: Perception and Practice"
author: "Nathan Holtrop"
format: 
  html:
    output-file: "index"
    theme: lux
    toc: true
    code-fold: true
editor: 
  markdown: 
    wrap: 72
---

[← Back to Home](../index.html){.btn .btn-outline-secondary .btn-sm
role="button"}

----------------

## Exercise 1: Wilke on visualizing amounts

Read Chapter 6 of Wilke (2019). This chapter begins by discussing bar charts. Many of you gravitate toward bar plots. Beware the tendency to overuse them! But if you are going to use them, you should use them well.

1. List some guidelines/advice Wilke gives about creating bar charts.

  - Always start at a zero baseline!
  
  - Use horizontal bars if the tables are too long.
  
  - Rearrange bars by size unless they are already in a meaningful order.

2. When is it not advised to use a bar chart? Why?

It is not advised to use a bar chart if the bars do not all start at zero or if the difference between the height of the bars is minute. This is not an efficient use of ink and it is difficult to decipher. It would likely be better to use something like a scatter plot.

3. What alternatives to bars are mentioned in this chapter?

  - Heatmap - "While this visualization makes it harder to determine the exact data values shown, it does an excellent job of highlighting broader trends."
  
  - Dot plot - Much better for location rather than length.

4. What guidance does Wilke give about whether or not to stack bars vs. dodge them (using an offset in Vega-Lite)?

Wilke suggests that stacking bars is useful when the sum of the bars is a meaningful amount or if the values are representing a count. You should not use stacked bars if the primary goal is to compare across subcategories as they will not all have a zero baseline.

5. Recreate Figure 6.3 in Vega-Lite.

Import the Vegawidget Library
```{r}
library(vegawidget)
```

``` {r}
'
{
  "$schema": "https://vega.github.io/schema/vega-lite/v5.json",
  "data": {
    "url": "https://calvin-data304.netlify.app/data/cow-movies.csv"
  },
  "width": 400,
  "height": 200,
  "mark": "bar",
  "encoding": {
    "x": {
      "field": "amount",
      "type": "quantitative",
      "title": "Weekend gross (USD)"
    },
    "y": {
      "field": "title",
      "type": "nominal",
      "sort": "-x",
      "title": null
    }
  }
}' |> as_vegaspec()
```

6. Recreate Figures 6.8 and 6.9 in Vega-Lite.

Figure 6.8

``` {r}
'
{
  "$schema": "https://vega.github.io/schema/vega-lite/v5.json",
  "data": {
    "url": "https://calvin-data304.netlify.app/data/cow-income.csv"
  },
  "transform": [
    {"filter": "datum.race != \'all\' && datum.race != \'white_non_hispanic\'"}
  ],
  "width": 400,
  "height": 300,
  "mark": "bar",
  "encoding": {
    "x": {
      "field": "race",
      "type": "nominal",
      "title": "race",
      "axis": {"labelAngle": 0}
    },
    "y": {
      "field": "median_income",
      "type": "quantitative",
      "title": "median income (USD)"
    },
    "color": {
      "field": "age",
      "type": "nominal",
      "title": "age (years)",
      "scale": {"scheme": "viridis"}
    },
    "xOffset": {"field": "age"}
  }
}' |> as_vegaspec()
```

Figure 6.9

``` {r}
'
{
  "$schema": "https://vega.github.io/schema/vega-lite/v5.json",
  "data": {
    "url": "https://calvin-data304.netlify.app/data/cow-income.csv"
  },
  "transform": [
    {
      "filter": "datum.race != \'all\' && datum.race != \'white_non_hispanic\'"
    }
  ],
  "facet": {
    "field": "race",
    "type": "nominal",
    "columns": 2,
    "title": null,
    "header": {"labelFontSize": 12, "labelFontWeight": "bold"}
  },
  "spec": {
    "width": {"step": 45},
    "height": 200,
    "mark": "bar",
    "encoding": {
      "x": {
        "field": "age",
        "type": "nominal",
        "title": "age (years)",
        "sort": ["15-24", "25-34", "35-44", "45-54", "55-64", "65-74", ">74"],
        "axis": {"labelAngle": 0}
      },
      "y": {
        "field": "median_income",
        "type": "quantitative",
        "title": "median income (USD)"
      },
      "color": {
        "field": "race",
        "type": "nominal",
        "legend": null
      }
    }
  },
  "resolve": {"scale": {"y": "shared", "x": "shared"}}
}' |> as_vegaspec()
```

7. Recreate Figure 6.11 and explain why Figures 6.12 and 6.13 are labeled “bad”. (Note: You will have to do some filtering or do something extra to deal with the extra data here.)

Figure 6.11

``` {r}

'
{
  "$schema": "https://vega.github.io/schema/vega-lite/v5.json",
  "data": {
    "url": "https://calvin-data304.netlify.app/data/cow-gapminder.csv"
  },
  "transform": [
    {"filter": "datum.continent == \'Americas\' && datum.year == 2007"}
  ],
  "width": 400,
  "height": 450,
  "mark": {"type": "point", "filled": true, "color": "#0072B2", "size": 60},
  "encoding": {
    "x": {
      "field": "lifeExp",
      "type": "quantitative",
      "title": "life expectancy (years)",
      "scale": {"zero": false}
    },
    "y": {
      "field": "country",
      "type": "nominal",
      "sort": "-x",
      "title": null
    }
  }
}' |> as_vegaspec()

```

Figure 6.12 is bad because the bars are too long and draw the eye away from the key difference of life expectancy that is being displayed which is the primary focus of the graph. Figure 6.13 is bad because the countries are ordered alphabetically, which causes a dots to become disordered and hectic.

----------------

## Exercise 2: A video presentation by Healy

1. Watch this video starting at 12:18 until 49:52. (You are welcome to watch the whole thing if you like.) You will see some things that look familiar from our discussion of perception in class, but there are also a number of additional examples and discussion.

2. List at least three pieces of advice you can glean from this video that might help you design good graphics.
    
    - A well-designed workhorse can do a lot: You don't need a spectacular graphic to accomplish your goals and do so effectively. 
    
    - Remember to highlight the parts of a graphic that you want viewers to notice and focus on.
    
    - Make sure that the design of the graphic matches the "context" of the data. Map variables clearly and avoid using aesthetic features that will distort the true representation.

3. There are two figures in this video that come from Chapter 1 of Tufte (2001). Did you spot them? Which figures are they?

  - The two figures that matched were Charles Joseph Minard’s Map of Napoleon’s Russia Campaign and, despite scouring the video as much as five times, I was unable to find the second one.

----------------

## Exercise 3: Heat maps

1. In Vega-Lite lingo, what makes something be a heat map?

Using Vega-Lite terminology, something is a heatmap if the graph is encoded to have two categorical variables on the x and y axes, with the rectangle mark, and a quantitative variable encoded to color.

2. Recreate Figure 6.14 or 6.15 from Wilke (2019) (your choice). Or get fancy and include an interactive element that let’s you select the year to order by.

``` {r}

'
{
  "$schema": "https://vega.github.io/schema/vega-lite/v5.json",
  "data": {
    "url": "https://calvin-data304.netlify.app/data/cow-internet2.csv",
    "format": {"type": "csv", "parse": {"users": "number"}}
  },
  "transform": [
    {"filter": "isValid(datum.users)"}
  ],
  "width": 600,
  "height": 450,
  "mark": {"type": "rect", "stroke": "white", "strokeWidth": 0.5},
  "encoding": {
    "x": {
      "field": "year",
      "type": "ordinal",
      "title": "Year",
      "axis": {"labelAngle": -45}
    },
    "y": {
      "field": "country",
      "type": "nominal",
      "title": null,
      "sort": {"field": "users", "op": "max", "order": "descending"}
    },
    "color": {
      "field": "users",
      "type": "quantitative",
      "title": "Internet Users (%)",
      "scale": {"scheme": "magma"}
    }
  },
  "config": {
    "view": {"stroke": "transparent"},
    "axis": {"domain": false, "ticks": false}
  }
}' |> as_vegaspec()
```



3. At around 20:10, Healy presents a possible problem with heat maps, what is it?

Healy explains that, while we can tell general differences between colors, color is an incredibly innaccurate way of comparing continuous values. Firstly, because it is difficult to compare such similar values, and, secondly, because of the effect that the surrounding colors have on the square we are focused on. The surrounding colors will make it seem either lighter or darker than it actually is.

4. But starting at around 44:12, he presents a heat map as a “show pony”. Why isn’t the problem presented earlier an issue here? Is it an issue in the figure you make in part b? Why or why not?

The problem mentioned previously does not matter with the "show pony" due to the fact that we are looking less at the individual pixels and far more at the general trends and major changes over time.

----------------

## Exercise 4: Pie charts

1. What does Healy have to say about pie charts in his video?

Healy explains that pie charts are much more difficult to interpret because we are far better at decoding lengths (bar charts) and position (scatter plots) as opposed to the relative angles and areas of a pie chart.

2. How does that compare to what Wilke says in Section 10.1 of Wilke (2019)?

Wilke's opinions seem to contradict with Healy as he claims that there are certain use cases where pie charts are favorable. He says that pie charts work well when the goal is to "emphasize simple fractions" or when datasets are quite small.

3. What alternatives does Wilke present to pie charts and in what situations does he favor each? (You should be comparing pie charts to two other types of graphics that can be made with the same sort of data.)

WIlke compares pie charts to both stacked bar chart and side-by-side bar charts. He says that similar to pie charts, stacked bar charts do well to visualizes the data as proportions of a whole. He then continues by explaining that side-by-side bars allow for the easiest visualization of relative proportions and work well when the whole is broken into many pieces. Similarly to pie charts, side-by-side bars work great for smaller datasets but that stacked bars are the only ones that work well as a visualization of many sets of proportions or time series of proportions.

4. We won’t make many pie charts, but I suppose you should make at least one. To do this in Vega-Lite, use an arc mark. Recreate Figure 10.1 of Wilke (2019).

``` {r}

'
{
  "$schema": "https://vega.github.io/schema/vega-lite/v5.json",
  "description": "Recreation of Wilke Figure 10.1: German Parliament Composition",
  "data": {
    "values": [
      {"party": "FDP", "seats": 39},
      {"party": "SPD", "seats": 214},
      {"party": "CDU/CSU", "seats": 243}
    ]
  },
  "width": 500,
  "height": 500,
  "encoding": {
    "theta": {"field": "seats", "type": "quantitative", "stack": true},
    "color": {
      "field": "party",
      "type": "nominal",
      "scale": {
        "domain": ["CDU/CSU", "SPD", "FDP"],
        "range": ["#a9a9a9", "#AA4A44", "#FFDB58"]
      },
      "legend": null
    },
    "order": {"field": "seats", "type": "quantitative", "sort": "descending"}
  },
  "layer": [
    {
      "mark": {"type": "arc", "outerRadius": 150, "stroke": "white", "strokeWidth": 2}
    },
    {
      "mark": {"type": "text", "radius": 185, "fontSize": 16, "fontWeight": "bold"},
      "encoding": {
        "text": {"field": "party", "type": "nominal"}
      }
    },
    {
      "mark": {"type": "text", "radius": 100, "fill": "white", "fontSize": 18, "fontWeight": "bold"},
      "encoding": {
        "text": {"field": "seats", "type": "quantitative"}
      }
    }
  ],
  "view": {"stroke": null}
}' |> as_vegaspec()

```

----------------

# References

Tufte, Edward R. 2001. The Visual Display of Quantitative Information. 2nd ed. Cheshire, CT: Graphics Press. https://kyl.neocities.org/books/%5BTEC%20TUF%5D%20the%20visual%20display%20of%20quantitative%20information.pdf.

Wilke, C. O. 2019. Fundamentals of Data Visualization: A Primer on Making Informative and Compelling Figures. O’Reilly Media. https://clauswilke.com/dataviz/.