---
title: "Homework 5: Data Visualization Analysis"
author: "Nathan Holtrop"
format: 
  html:
    output-file: "index"
    theme: lux
    toc: true
    code-fold: true
editor: 
  markdown: 
    wrap: 72
---

[â† Back to Home](../index.html){.btn .btn-outline-secondary .btn-sm
role="button"}

## Test

## Exercise 1: Linking

Complete!

## Exercise 2: Gallery Analysis

### The Graphic

![Ridgeline Plot of Temperatures in Lincoln, NE by month in
2016](images/ExampleVisualization.png){width="40%" fig-align="center"}

[Link to original
visualization](https://r-graph-gallery.com/294-basic-ridgeline-plot.html#color)

### Analysis

-   **Marks & Channels:**
    -   **Marks:** Line and Area
    -   **Mapping:**
        -   Month is mapped to the y-axis with faceted x-axes for each
            month.
        -   Mean Temperature \[F\] is mapped to the x-axis.
            -   The height of each ridgeline is determined by the
                density of that temperature occurring.
        -   Color (a continuous representation of Mean Temperature
            \[F\]) is mapped to fill.
-   **The Story: What is this graphic trying to tell the viewer?**

This graph not only demonstrates the temperature in Lincoln, NE every
month, but also highlights the variance of the temperature. This is
important for discussing the effects of seasonality. For instance, the
winter months (i.e. December and January) show not only colder
temperatures, but also increased variance with shorter and longer plots.
This contrasts with the summer months of June and July which have much
more peaky and narrow graphs, emphasizing the fairly consistent and
warmer temperatures.

-   **Why it works: Why did you pick this? What makes it "good"?**

I picked this graphic because it stuck out to me as an interesting
choice of plotting in addition to internal faceting (they would likely
be referred to as something else within Vega-Lite). I found it fairly
simple to read and interpret while also exhibiting the strength of
conveying a large amount of information. Furthermore, I think the graph
made good use of the color scheme, using warmer colors for warmer
temperatures and cooler colors for the winter months.

In addition to all that is mentioned above. I appreciated the smooth
ridges in comparison to another option such as a histogram because it
demonstrates the continuity of time.

-   **Vega-Lite What features would you know how to implement in
    Vega-Lite? What features would you not know how to do (or are
    tricky) in Vega-Lite?**

Within Vega-Lite, and with the training that we have achieved thus far,
I am confident that I could implement the 'Area' mark, map variables to
the x and y axes, create some form of a ridgeplot, implement a variation
of the facets, and add a title and labels.

Something that might be tricky to do would be to generate the smooth
curves on the plots from raw data points and generate the proper facets
required. The biggest issue would likely be the overlap between the
ridgeline plots for each month.

------------------------------------------------------------------------
## Exercise 3: Graph-Making Exercises (Seattle Weather)

Import the Vegawidget Library
```{r}
library(vegawidget)
```

### Exercise #1
```{r}
'
{
  "$schema": "https://vega.github.io/schema/vega-lite/v5.json",
  "description": "Daily High Temperature in Seattle",
  "data": {
    "url": "https://calvin-data304.netlify.app/data/weather-with-dates.csv"
  },
  "width": 600,
  "height": 300,
  "mark": "point",
  "encoding": {
    "x": {
      "field": "date",
      "type": "temporal",
      "timeUnit": "dayofyear",
      "title": "Day of the Year"
    },
    "y": {
      "field": "temp_max",
      "type": "quantitative",
      "title": "Max Temperature (C)"
    },
    "color": {
      "field": "year",
      "type": "nominal",
      "title": "Year"
    },
    "opacity": {
      "value": 0.5
    }
  }
}' |> as_vegaspec()
```

### Exercise #3
```{r}
'
{
  "$schema": "https://vega.github.io/schema/vega-lite/v5.json",
  "description": "Distribution of Weather Types by Month",
  "data": {
    "url": "https://calvin-data304.netlify.app/data/weather-with-dates.csv"
  },
  "width": 600,
  "height": 300,
  "mark": "bar",
  "encoding": {
    "x": {
      "field": "date",
      "type": "ordinal",
      "timeUnit": "month",
      "title": "Month"
    },
    "y": {
      "aggregate": "count",
      "type": "quantitative",
      "title": "Number of Days"
    },
    "color": {
      "field": "weather",
      "type": "nominal",
      "title": "Weather Condition",
      "scale": {
        "domain": ["sun", "fog", "drizzle", "rain", "snow"],
        "range": ["#e7ba52", "#a7a7a7", "#aec7e8", "#1f77b4", "#9467bd"]
      }
    }
  }
}' |> as_vegaspec()
```