---
title: "Final Portfolio"
author: "Nathan Holtrop"
format: 
  html:
    output-file: "index"
    theme: lux
    toc: true
    code-fold: true
editor: 
  markdown: 
    wrap: 72
---

[← Back to Home](../index.html){.btn .btn-outline-secondary .btn-sm
role="button"}

----------------

## Exercise 1: HW #4 Revision

1. Scroll through the HW 4 Gallery to see the plots we created at that time. Find an example that has that has something you like about it and explain what you like.

    - One of my favourite graphs from the gallery was Example 2B "Comparing Kits". I found the main goal of the graph, kit comparison, especially easy to interpret (once the mapping of the color scheme to region was memorized) due to the distinctly clear comparison of genetic share proportion by kit. For instance, for the first set of twins, it was clear that Ancestry "favored" NW Europe over SE Europe or potentially used different geographical definitions.

2. Find an example that has something you don’t like, and explain what you don’t like about it.

    - I was unable to see the full value of Example #4B. While I appreciate the attempted use of different shapes, there was a lot of overlap on the graph and a complete lack of notation regarding geographics. While I understand that the point of the graph was to instead compare the similarities/differences between twins, I feel as though the geographical data was lost and the scale of the facets was too small to see the minute differences between twins.

3. Now create two graphics, one that helps compare the kits and one that helps compare twins. Give your graphics good titles, use use other principles of good visualization, and do not restrict your plot to just a small subset of the data.

```{r}
library(vegawidget)
```

```{r}

'
{
  "$schema": "https://vega.github.io/schema/vega-lite/v5.json",
  "title": {
    "text": "DNA Kit Profile Comparison",
    "subtitle": "How different laboratories interpret the same 12 individuals",
    "anchor": "start",
    "fontSize": 18
  },
  "data": {
    "url": "https://calvin-data304.netlify.app/data/twins-genetics-long.json"
  },
  "width": 600,
  "height": 100,
  "spacing": 15,
  "facet": {
    "row": {
      "field": "kit",
      "type": "nominal",
      "title": null,
      "header": {"labelFontSize": 12, "labelFontWeight": "bold"}
    }
  },
  "spec": {
    "mark": {"type": "bar", "tooltip": true},
    "encoding": {
      "x": {
        "field": "twin",
        "type": "nominal",
        "title": "Individual Participant",
        "axis": {"labelAngle": 0}
      },
      "y": {
        "field": "genetic share",
        "type": "quantitative",
        "title": "Genetic Share",
        "axis": {"format": "%"}
      },
      "color": {
        "field": "region",
        "type": "nominal",
        "scale": {"scheme": "tableau10"},
        "title": "World Region"
      }
    }
  },
  "config": {
    "view": {"stroke": "transparent"}
  }
}' |> as_vegaspec()

```

This visualization highlights the inherent subjectivity in genetic testing. By aligning the results of Ancestry, 23andMe, and MyHeritage into three distinct rows, we can see that the "DNA story" for a single person (such as 1A or 4B) changes depending on the kit they use. For example, some kits are much more aggressive in identifying specific regions like "SE Europe," while others lump that same genetic material into "NW Europe." The visual consistency of the bars across the X-axis for each individual demonstrates that while the underlying DNA is constant, the algorithms used by these labs create varying interpretations of one’s heritage.

```{r}
'
{
  "$schema": "https://vega.github.io/schema/vega-lite/v5.json",
  "title": {
    "text": "Twin Genetic Consistency Analysis",
    "subtitle": "Comparing identical twin pairs (A & B) across three DNA kits",
    "anchor": "start",
    "fontSize": 18
  },
  "data": {
    "url": "https://calvin-data304.netlify.app/data/twins-genetics-long.json"
  },
  "width": 180,
  "height": 250,
  "facet": {
    "column": {
      "field": "kit",
      "type": "nominal",
      "title": "Testing Kit",
      "header": {"labelFontSize": 14}
    }
  },
  "spec": {
    "mark": {"type": "bar", "stroke": "white", "strokeWidth": 0.5},
    "encoding": {
      "x": {
        "field": "pair",
        "type": "nominal",
        "title": "Twin Pair Number"
      },
      "xOffset": {"field": "id"},
      "y": {
        "field": "genetic share",
        "type": "quantitative",
        "title": "Genetic Share",
        "axis": {"format": "%"}
      },
      "color": {
        "field": "region",
        "type": "nominal",
        "title": "Region",
        "scale": {"scheme": "tableau10"}
      },
      "tooltip": [
        {"field": "twin", "title": "Person"},
        {"field": "region", "title": "Region"},
        {"field": "genetic share", "format": ".1%", "title": "Share"}
      ]
    }
  },
  "config": {
    "view": {"stroke": "transparent"}
  }
}' |> as_vegaspec()

```

Because identical twins share 100% of their DNA, their ancestry results should, in theory, be identical. This graphic tests that theory by pairing twins 1A/1B through 6A/6B side-by-side. The results show a surprising lack of consistency: even within the same company (like MyHeritage), twins in Pair 1 and Pair 4 received noticeably different regional estimates. The white borders between the bars help to emphasize these "steps" in height where the twins' data diverges. This visualization reveals that the minute variations in how DNA is sampled and processed can lead to different results for the exact same genetic code, posing a challenge to the perceived precision of consumer DNA kits.

4. For each graphic, include a paragraph that tells the story of your graphic.

    - See above

## Exercise 2: Data and Graphics Challenge

## Exercise 3: A New Challenge

1. Enter the data into Excel, a CSV, or JSON file. You will have some decisions to make about things like variable names, etc. Be sure to include a link to the data set you create on your portfolio website. Notice that some of the surveys were conducted all in one year and some spanned two calendar years. How will you deal with that?

2. Use these data to create a visualization that tells a story.

3. Write a few sentences explaining the story told.

4. Technical notes.

## Exercise #4: My Masterpiece

1. Using a data set of your choosing, create a graphic that demonstrates your abilities to design and create a graphic that tells a compelling story.

2. Explain the choices you made when designing your graphic and relate them to principles of good graphics that we have learned or seen in this class. Mention alternatives to your graphic that you considered but did not opt to submit.

3. Be sure to include information about where you got your data from. This could be a link to a website, a proper citation of an article, a description of a research project you have been working on, etc.

## Exercise #5: Using your palette

1. The grammar of graphics gives us a palette of graphical elements with which to “paint” our graphic. The palette includes various marks, channels, composition, etc. One of the goals for your portfolio is that you demonstrate the ability to use a variety of these features and use them effectively. Look over your graphics, and identify a place where you used:

    a. an encoding channel other than x or y.
    b. layers
    c. facets
    d. concatenation or repeat
    e. non-default settings for a channel’s scale or guide
    f. tooltips
    g. another kind of interaction (panning/zooming, brushing, sliders, etc.)

## Exercise #6: Using your palette

1. Cite 2 or 3 specific examples in the graphics in your portfolio where you used a feature of Vega-Lite/vegabrite/Altair/altair that we did not learn in class. This might be using a new kind of mark or transform, or a way to customize a feature of the graphic, or a way to use interaction, or…

2. You will also want to keep learning about principles of good graphics design. Cite 2 or 3 specific examples where you followed the advice in one of the resources from class. Provide specific page (for print or pdf) or section (for HTML) references. Include direct links, if possible. (HTML books often make it easy to link directly to a section.)

## References
Healy, K. 2019. Data Visualization: A Practical Introduction. Princeton University Press. https://socviz.co/.
Heer, Dominik, Jeffrey AND Moritz. 2024. “Mosaic: An Architecture for Scalable & Interoperable Data Views.” IEEE Trans. Visualization & Comp. Graphics (Proc. VIS). https://doi.org/10.1109/TVCG.2023.3327189.
Knaflic, C. N. 2015. Storytelling with Data: A Data Visualization Guide for Business Professionals. Wiley. https://github.com/Saurav6789/Books-/blob/master/storytelling-with-data-cole-nussbaumer-knaflic.pdf.
———. 2020. Storytelling with Data: Let’s Practice! Wiley. https://github.com/Saurav6789/Books-/blob/master/Storytelling%20with%20Data%20Let%
E2%80%99s%20Practice%20by%20Cole%20Nussbaumer%20Knaflic%20(z-lib.org).pdf.
Tufte, Edward R. 2001. The Visual Display of Quantitative Information. 2nd ed. Cheshire, CT: Graphics Press. https://kyl.neocities.org/books/%5BTEC%20TUF%5D%20the%20visual%20display%20of%
20quantitative%20information.pdf.
Wilke, C. O. 2019. Fundamentals of Data Visualization: A Primer on Making Informative and Compelling Figures. O’Reilly Media. https://clauswilke.com/dataviz/.

