---
title: "Annual Ticket Performance Report"
subtitle: "How the team mastered seasonal surges to deliver record service levels"
author: Nathan Holtrop
format: 
  revealjs:
    theme: simple
    transition: slide
    incremental: true 
    code-fold: true
---

## Executive Summary

This year was defined by two significant periods of seasonal pressure in **May** and **November**. Despite these inventory spikes, the team achieved a landmark turnaround in service quality, ending the year significantly above our 90% performance goal.

---

## 1. The Pressure: Seasonal Inventory Spikes

To understand our year, we must look at the "Work-in-Process" (WIP). Our inventory levels are not static; they are heavily influenced by seasonal demand.

---

```{r}

library(vegawidget)

'
{
  "$schema": "https://vega.github.io/schema/vega-lite/v5.json",
  "data": {"url": "knaflic-6-11.csv"},
  "width": 800,
  "height": 450,
  "title": {
    "text": "Total Work-in-Progress vs. Pending eFile",
    "subtitle": "Combined volume across all channels compared to the eFile bottleneck",
    "anchor": "start",
    "fontSize": 24,
    "subtitleFontSize": 16
  },
  "transform": [
    {
      "calculate": "toNumber(datum.Requests) + toNumber(datum.Orders) + toNumber(datum[\'Pending eFile\']) + toNumber(datum[\'Unprocessed Inbox\'])",
      "as": "TotalVolume"
    }
  ],
  "layer": [
    {
      "mark": {"type": "area", "opacity": 0.1, "color": "gray"},
      "encoding": {
        "x": {
          "field": "Month", 
          "type": "nominal", 
          "sort": null,
          "axis": {
            "labelAngle": 0, 
            "titlePadding": 15,
            "labelFontSize": 14,
            "titleFontSize": 16
          }
        },
        "y": {
          "field": "TotalVolume", 
          "type": "quantitative", 
          "title": "Ticket Volume",
          "axis": {
            "labelFontSize": 14,
            "titleFontSize": 16
          }
        }
      }
    },
    {
      "mark": {"type": "line", "point": true, "color": "black", "strokeDash": [4,2], "size": 2},
      "encoding": {
        "x": {"field": "Month", "type": "nominal", "sort": null},
        "y": {"field": "TotalVolume", "type": "quantitative"}
      }
    },
    {
      "mark": {"type": "line", "point": {"size": 60}, "color": "#ff7300", "size": 4},
      "encoding": {
        "x": {"field": "Month", "type": "nominal", "sort": null},
        "y": {"field": "Pending eFile", "type": "quantitative"}
      }
    },
    {
      "transform": [{"filter": "datum.Month == \'DEC\'"}],
      "encoding": {
        "x": {"field": "Month", "type": "nominal", "sort": null},
        "y": {"field": "TotalVolume", "type": "quantitative"}
      },
      "layer": [
        {
          "mark": {
            "type": "text", 
            "align": "left", 
            "dx": 15, 
            "dy": -10, 
            "fontWeight": "bold",
            "fontSize": 14
          },
          "encoding": {
            "text": {"value": "TOTAL VOLUME"},
            "color": {"value": "black"}
          }
        },
        {
          "mark": {
            "type": "text", 
            "align": "left", 
            "dx": 15, 
            "dy": 15, 
            "fontWeight": "bold",
            "fontSize": 14
          },
          "encoding": {
            "y": {"field": "Pending eFile", "type": "quantitative"},
            "text": {"value": "PENDING EFILE"},
            "color": {"value": "#ff7300"}
          }
        }
      ]
    }
  ],
  "config": {
    "view": {"stroke": "transparent"},
    "autosize": {"type": "pad", "resize": true}
  }
}
' |> as_vegaspec()

```

---

**Key Takeaway:** The Work-in-Progress Inventory level spiked significantly in both May and November. This was driven primarily by (or clearly correlated with) sudden jumps in the "Pending eFile," creating a backlog leading into the summer and Christmas seasons.

---

## 2. Front-Office Consistency
While the back office dealt with inventory surges, the front office remained remarkably stable. This metric is crucial because it proves that they did not sacrifice customer access even when internal backlogs were high.

---

```{r}
#| label: reachability-performance-plot
#| warning: false
#| message: false
#| fig-width: 9
#| fig-height: 5

library(ggplot2)
library(readr)
library(scales)
library(dplyr)

reachability_data <- read_csv("knaflic-6-11.csv") %>%
  mutate(
    Reach_Value = parse_number(as.character(`Front Office Reachability`)),
    Reach_Value = ifelse(Reach_Value > 1, Reach_Value / 100, Reach_Value),
    Month = factor(Month, levels = unique(Month))
  )

# 2. Create the Bar Chart
ggplot(reachability_data, aes(x = Month, y = Reach_Value)) +
  geom_col(fill = "#90EE90", alpha = 0.8) + 
  geom_hline(yintercept = 0.90, linetype = "dashed", color = "red", size = 1.2) +
  geom_text(aes(label = percent(Reach_Value, accuracy = 0.1)), 
            vjust = -0.5, size = 4, fontface = "bold") +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1.1)) +
  labs(
    title = "Front Office Reachability: Monthly Consistency",
    subtitle = "Performance remained near-perfect (97%+) regardless of back-office volume",
    x = NULL, y = "Calls Reached (%)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(face = "bold")
  ) +
  # Goal Label
  annotate("text", x = 1.5, y = 0.85, label = "GOAL: 90%", color = "red", fontface = "bold")

```

---

**Key Takeaway:** Throughout the year, Front Office Reachability stayed remarkably high (consistently above 97%). While average answer times hovered slightly above their 8-second goal, the lines of communication remained open and reliable regardless of volume.

---

## 3. Scaling the Standard Service Levels
The focus metric of the year is their Service Level Agreement (SLA) percentage. They moved from a position of failure in January to a position of mastery by December.

---

```{r}
#| label: sla-performance-final-fix
#| warning: false
#| message: false

library(ggplot2)
library(readr)
library(scales)
library(dplyr)

ticket_data <- read_csv("knaflic-6-11.csv") %>%
  mutate(
    SLA_Value = parse_number(as.character(`Overall Result`)),
    SLA_Value = ifelse(SLA_Value > 1, SLA_Value / 100, SLA_Value),
  
    Month = factor(Month, levels = unique(Month))
  )
ggplot(ticket_data, aes(x = Month, y = SLA_Value)) +
  geom_col(fill = "#4682b4", alpha = 0.9) + 
  geom_hline(yintercept = 0.90, linetype = "dashed", color = "red", size = 1.2) +
  geom_text(aes(label = percent(SLA_Value, accuracy = 1)), 
            vjust = -0.5, size = 4.5, fontface = "bold") +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1.1)) +
  labs(
    title = "Monthly Service Level Performance (SLA %)",
    subtitle = "Consistent growth from 82% in January to 97% by December",
    x = NULL, y = "SLA Achievement Rate"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(face = "bold")
  ) +
  annotate("text", x = 1.5, y = 0.94, label = "GOAL: 90%", color = "red", fontface = "bold")

```

---

**Key Takeaway:** The organization successfully transformed its delivery model. Starting from just 82% in January, they steadily increased their efficiency, easily surpassing the 90% goal and ending the year at a record 97% success rate.

---

## Conclusion

The data confirms that the team is now better equipped to handle seasonality than it was twelve months ago. Even during the massive November inventory spike, they maintained higher service levels than they did during the smaller May spike, a clear indicator of operational maturity.